---
title: "speeding st_segments up with data.table"
output: html_notebook
---

Although I  already saw a drastic performance improvement with [`st_segments_tbl`](), I still wanted to go ahead and work with `data.table` to see if the function can be made faster and more efficient.

```{r getting a road network, message=F, warning=F,echo =F}
library(sf)
library(dplyr)
library(tigris)

midd = read_sf("https://opendata.arcgis.com/datasets/625b92329bb841cf8295d14553b67e1f_2.geojson") %>% 
  filter(COLLEGE == "MIDDLEBURY COLLEGE") %>% 
  st_transform(32145) %>% # Vermont State Plane Meters
  st_union() %>% 
  st_centroid() %>% 
  st_buffer(5000) %>% 
  st_bbox %>% 
  st_as_sfc

streets = roads("vt", "addison county", progress_bar = F) %>% st_transform(32145) %>% st_intersection(midd) %>%
  filter(!grepl("MULTI", as.character(st_geometry_type(.))))
```

```{r st_segments_dt}
st_segments_dt = function(x, max_length) {
  # breaks linestrings/polygons into line segments, keeps attributes
  # x: layer with linestrings
  # max_length: optional parameter, max length of segments, unit from crs
  library(sf)
  library(data.table)
  library(tibble)
  
  x$id = 1:nrow(x)
  
  if (!missing(max_length)) {
    x = st_segmentize(x, max_length)
  }
  
  coord = st_coordinates(x)
  col = ncol(coord)
  colname = colnames(coord)[col]
  
  geom = as.data.table(coord)
  geom[, `:=`(x2 = shift(X, type = "lead"), y2 = shift(Y, type = "lead")), by = colname]
  geom = na.omit(geom)
  geom[, rows := .I]
  geom[,geometry := .(list(st_linestring(rbind(c(X,Y),c(x2,y2))))), by = rows]
  geom = st_sfc(geom$geometry, crs = st_crs(x))
  
  fid = as.data.table(coord[, col])
  colnames(fid) = "id"
  setkey(fid, id)
  fid = fid[fid[, .I[-1], by = id]$V1][as.data.table(st_drop_geometry(x)), on = "id"]
  fid[, `:=`(geometry = geom, id = NULL)]
  st_sf(as_tibble(fid))
}
```

`st_segments_dt` essentially has the same structure as `st_segments_tbl`, though everything save for the last line with `as_tibble` is done using either `data.table` or `sf`. `as_tibble` is not entirely necessary here but I prefer the print method of tibbles over data.tables and `sf` supports tibbles.

```{r dplyr, echo = F}
st_segments_tbl = function(x, max_length) {
  library(sf)
  library(dplyr)
  library(purrr)
  
  layer = mutate(x, value = row_number())
  
  if (!missing(max_length)) {
    layer = st_segmentize(layer, max_length)
  }
  
  coord = st_coordinates(layer)
  col = ncol(coord)
  
  geom = as_tibble(coord) %>%
    group_by(!!sym(colnames(coord)[col])) %>%
    mutate(X2 = lead(X),
           Y2 = lead(Y)) %>%
    ungroup() %>%
    filter(!is.na(X2)) %>%
    mutate(geometry = pmap(list(X, Y, X2, Y2), function(x, y, x2, y2)
      st_linestring(rbind(
        c(x, y), c(x2, y2)
      )))) %>%
    {
      st_sfc(.$geometry, crs = st_crs(layer))
    }
  
  as_tibble(coord[, col]) %>%
    group_by(value) %>%
    slice(-1) %>%
    ungroup %>%
    left_join(st_drop_geometry(layer), by = "value") %>%
    select(-1) %>%
    mutate(geometry = geom) %>%
    st_sf()
}
```

```{r benchmark, message = F, warning=F}
bench::mark(
  "dplyr/purrr" = st_geometry(st_segments_tbl(streets,1)),
  data.table = st_geometry(st_segments_dt(streets,1)),
  iterations = 10
) %>%
  mutate(name = as.character(expression),
         max = bench::as_bench_time(lapply(time, max))) %>%
  select(
    name,
    min,
    median,
    max,
    "memory allocation" = mem_alloc,
    iterations = n_itr,
    "total time" = total_time
  )
```
 `st_segments_dt` provides a slight improvement in speed and is more efficient than `st_segments_tbl`. Both functions are still considerably faster than `st_segments`. 