---
title: "fixing pg_write_stars"
output: html_notebook
---

I found the cause of the shifting cells/pixels in some of the raster tiles was due an a function in `DBI`. When using `dbWriteTable` to write the values of each raster tile to the database, the table became unordered in some cases. 

```{r examples, message = F,warnings= F}
library(stars)
library(dplyr)
library(keyring)
library(DBI)

flood = read_stars("floodextent200.tif")

xy = dim(flood)

values = tibble(val = c(flood[[1]]), x = sort(rep(1:xy[2],xy[1])))

conn = DBI::dbConnect(
  RPostgres::Postgres(),
  dbname = key_list()[1, 1],
  user = key_list()[1, 2],
  password = key_get(key_list()[1, 1], key_list()[1, 2])
)

dbWriteTable(conn,"value_test",values, overwrite =T)

write_test = dbReadTable(conn,"value_test")

dbRemoveTable(conn,"value_test")

sum(write_test$x != values$x)

cbind(write_test,values)[write_test$x != values$x,]
```
Here there are at least 70 rows in the written table where the values do not match the original table, resulting in the shifting of values. ~~I believe this may be due to the many NAs in the value column.~~ This is not due to the NAs in the column as I had originally thought, but is by [design](https://github.com/r-dbi/odbc/issues/242).

```{r}
rm_na = flood

rm_na[is.na(rm_na)] = 0

zeros = tibble(val = c(rm_na[[1]]), x = sort(rep(1:xy[2],xy[1])))

dbWriteTable(conn,"value_test",zeros, overwrite = T)

write_test = dbReadTable(conn,"value_test")

dbRemoveTable(conn,"value_test")

cbind(write_test,values)[write_test$x != values$x,]
```
Removing the NAs from the raster before writing it to the database seems to solve the problem, though I think it would be best to refrain from altering the raster in this way. A better solution would be to add a column with row numbers to the values table and use this column to order the table after it is written to the database. I used this simple solution to fix `pg_write_stars`. 

```{r fixed function}
pg_write_stars =  function(conn, name, layer, tile_size = NULL,bit_depth = NULL) {
  # writes raster (stars) to a postgis database
  # conn: DBIConnection object
  # name: name of table to hold raster
  # layer: stars object 
  # tile_size: size of tiles to split raster into specified as vector with length 1 or 2, expressed as pixel width by height 
  # bit_depth: bit depth of raster
  
  library(stars)
  library(tibble)
  library(glue)
  library(purrr)
  library(DBI)
  
  if (!("postgis_raster" %in% dbGetQuery(conn, "select * FROM pg_available_extensions;")$name))
    stop("PostGIS extension not available.")
  
  if (any(class(layer) != "stars")) 
    stop("Object should be of class stars.")

  crs = st_crs(layer)$epsg
  if (is.na(crs)) {crs = 0; warning("SRID set to 0")}
  
  dims = st_dimensions(layer)
  res = c(dims$x$delta, dims$y$delta)
  bbox = st_bbox(layer)
  val = c(dims$x$values,dims$y$values)
  aff = attr(dims,"raster")$affine
  
  if (attr(dims, "raster")$curvilinear |
      any(!is.null(val)) |
      any(is.na(res))) {
    stop("Raster must be regular.")
  }
  
  # determine bit depth 
  # from https://github.com/mablab/rpostgis/blob/master/R/pgWriteRast.R 
  if (is.null(bit_depth)) {
    if (is.integer(layer[[1]])) {
      if (min(layer[[1]], na.rm = TRUE) >= 0) {
        bit_depth = "32BUI"
      } else {
        bit_depth = "32BSI"
      }
    } else {
      bit_depth = "32BF"
    }
  }
  
  # determine tile size
  if (!is.null(tile_size)) {
    if (length(tile_size) == 2) {
      bx = tile_size[1]
      by = tile_size[2]
    } else {
      bx = by = tile_size[1]
    }
  } else {
    bx = by = 100
  }
  
  to = dim(layer)
  
  x1 = seq(1,to[1],bx)
  if (x1[length(x1)] != to[1]) x2 = c(x1[-1]-1,to[[1]])
  width = map2(x1,x2,~c(.x,.y)) %>% do.call(what=rbind)
 
  y1 = seq(1,to[2],by)
  if (y1[length(y1)] != to[2]) y2 = c(y1[-1]-1,to[[2]])
  height = map2(y1,y2,~c(.x,.y))
  
  tiles = map(height,~cbind(width,.x[1],.x[2])) %>% do.call(what=rbind)
  
  
  # create a list with tile dimensions and values 
  info = map(1:nrow(tiles), function(x) {
    tile = tiles[x,]
    tile = layer[1,tile[1]:tile[2],tile[3]:tile[4]]
    list(dim(tile), st_bbox(tile)[c("xmin","ymax")], as.vector(tile[[1]]))
  })
  
  dbExecute(conn, glue("drop table if exists {name}"))
  dbExecute(conn, glue("create table {name}(rid serial primary key, rast raster)"))
  
  for (i in seq_along(info)) {
    xy = info[[i]][[1]]
    # create table with each value assigned to a row 
    dbWriteTable(conn,
                 "rst_tile_val",
                 tibble(
                   val = info[[i]][[3]],
                   x = sort(rep(1:xy[2], xy[1])),
                   row_n = 1:prod(xy)
                 ) ,
                 overwrite = T)
    
    offset = info[[i]][[2]]
    
    # use tile dimensions to create empty raster and populate raster with values from rst_tile_val
    dbExecute(conn, glue(
    "insert into {name}(rid, rast) 
     values({i},
     st_setvalues(
        st_addband(
            st_makeemptyraster({xy[1]}, {xy[2]}, {offset['xmin']},{offset['ymax']},{res[1]},{res[2]},{aff[1]},{aff[2]},{crs}),
            array[row(1,'{bit_depth}',0,-99999)]::addbandarg[]),
        1, 1, 1,(select array(select array_agg(val)from (select val, x from rst_tile_val order by row_n) as rst group by x order by x))::double precision[][]
     ))"))
  }
  dbExecute(conn,"drop table if exists rst_tile_val") 
  dbExecute(conn, glue("drop index if exists {name}_rast_idx"))
  # create spatial index
  dbExecute(conn, glue("create index {name}_rast_idx on {name} using gist(st_convexhull(rast))")) 
}
```

To implement the ordering solution in `pg_write_stars`, I had to first add a column with row numbers to the table of values when they are being written to the database.
```{r, eval = F}
dbWriteTable(con,
             "rst_tile_val",
             tibble(
               val = info[[i]][[3]], #tile values
               x = sort(rep(1:xy[2], xy[1])), # raster row number
               row_n = 1:prod(xy) # row number (new column)
             ) ,
             overwrite = T)
```

I then had to edit the SQL for inputting into an empty raster by adding a subquery to order the values by the new column with row numbers. This ensured that values of each column were ordered correctly before they were inputted in the raster. 

```{sql, eval =F}
select array(
    select array_agg(val) 
    from (
        select val, x from rst_tile_val order by row_n) as rst
        group by x order by x
      ) 
```

I also changed how the tiles were created, using a matrix of indices to subset the raster rather relying on intersections to break it up. I have yet to test whether this method is faster, though I like it more. 
```{r creating thet tiles, eval=F}
  # creating a matrix with indices for subsetting 
  x1 = seq(1,to[1],bx)
  if (x1[length(x1)] != to[1]) x2 = c(x1[-1]-1,to[[1]])
  width = map2(x1,x2,~c(.x,.y)) %>% do.call(what=rbind)
 
  y1 = seq(1,to[2],by)
  if (y1[length(y1)] != to[2]) y2 = c(y1[-1]-1,to[[2]])
  height = map2(y1,y2,~c(.x,.y))
  
  tiles = map(height,~cbind(width,.x[1],.x[2])) %>% do.call(what=rbind)
  
  # creating a list with tile dimensions and values 
  info = map(1:nrow(tiles), function(x) {
    tile = tiles[x,]
    tile = layer[1,tile[1]:tile[2],tile[3]:tile[4]]
    list(dim(tile), st_bbox(tile)[c("xmin","ymax")], as.vector(tile[[1]]))
  })
```

```{r pg_read_stars, echo=F}
pg_read_stars = function(conn, layer, rast = "rast"){
  # reads raster from postgis database
  # conn: DBIConnection object
  # layer: name of table with raster 
  # rast: name of column that contains raster 
  
  library(stars)
  library(purrr)
  library(DBI)
  library(glue)
  
  values = dbGetQuery(
    conn,
    glue(
      "with ordered as (
          select 
          {rast} as rast, 
          rid from {layer} 
          order by rid asc)
       select
       unnest(st_dumpvalues(rast,1)) as values,
       row_number() over (order by rid) as id
       from ordered"
    )
  )
  
  srid = dbGetQuery(
    conn,
    glue(
      "select distinct 
       st_srid({rast}) as srid 
       from {layer}"))$srid %>% 
    st_crs
  
  boxes = dbGetQuery(
    conn, 
    glue(
      "select 
       st_asewkb(st_envelope({rast})) as geom 
       from {layer}
       order by rid asc"))$geom %>% 
    st_as_sfc(EWKB = T) 
  
  dims = dbGetQuery(
    conn, 
    glue(
      "select  
       st_pixelwidth({rast}) as dx, 
       st_pixelheight({rast}) as dy,
       st_skewx({rast}) as sx,
       st_skewy({rast}) as sy
       from {layer}
       limit 1"))
  
  result = map2(boxes, unique(values$id), function(x, y)
    st_as_stars(
      st_bbox(x),
      dx = dims$dx,
      dy = dims$dy,
      values = values[values$id == y, ]$values
    ) %>% st_set_crs(srid)) %>%
    do.call(what = st_mosaic)
  
  attr(attr(result, "dimensions"), "raster")$affine = c(dims[[3]],dims[[4]])
  
  result
}
```

```{r reading and writing, message=F, warning=F, fig.height=6, fig.width=9}
library(mapview)


initial = flood

pg_write_stars(conn,"pg_stars_test", initial)

postgis = pg_read_stars(conn,"pg_stars_test")

sum((initial == postgis)[[1]],na.rm=T) == sum(!is.na(initial)[[1]])

mapview(initial, col.regions = "#EA3546", layer.name = "Initial Raster") +mapview(postgis, col.regions = "#226CE0", layer.name = "PostGIS Raster") 
```

Unlike with the first version of `pg_write_stars`, there now is no longer any random shifting and the raster and the rasters appear to be the same based on a visual inspection and a logical test. Because of this,I would consider this fix to be a success, albeit a small one. 
