---
title: "st_blend"
output: html_notebook
---

I needed a way to add nodes to a network when using pgRouting but did not have the desire to do so in PostGIS, so I wrote  a function which I can use in R before writing the layer to the database. 
```{r st_blend}
st_blend = function(node, network) {
  # blends node(s) into a network composed of linestrings
  # node: point(s)
  # network: network composed of linestrings
  # returns: network with added node(s)
  
  library(sf)
  library(lwgeom)
  library(dplyr)
  library(purrr)
  
  ncrs = st_crs(network)
  blade = st_sf(node) %>%
    mutate(geometry = st_geometry(.)) %>% # just in case geometry column isn't named geometry
    st_set_geometry("geometry") %>%
    mutate(
      nf = st_nearest_feature(., network),
      # nearest linestring
      nl = do.call(c, map2(
        geometry, nf,  ~ st_nearest_points(st_sfc(.x, crs = ncrs), network[.y, ])
      )),
      # nearest point to edge/linestring
      np = do.call(c, map2(
        nl, geometry, ~ st_cast(st_sfc(.x, crs = ncrs), "POINT") %>% .[. != st_sfc(.y, crs = ncrs)]
      )),
      # nearest point in nearest edge/linestring
      ns = do.call(c, map2(
        nf, np, ~ st_cast(st_geometry(network[.x, ]), "POINT") %>% .[st_nearest_feature(st_sfc(.y, crs = ncrs), .)]
      )),
      len = st_length(nl) %>% units::drop_units()
    ) %>%
    filter(np != ns) %>% # remove rows where nearest point is already in nearest edge/linestring
    st_set_geometry("nl") %>%
    select(nf, len) %>%
    suppressWarnings()
  
  blade = st_geometry(blade) %>%
    {
      (. - st_centroid(.)) * (blade$len * 2 / blade$len) + st_centroid(.) # ensure line is long enough to split edge
    } %>%
    st_set_crs(ncrs) %>%
    st_sf() %>%
    mutate(nf = blade$nf)
  
  st_split(network[blade$nf, ], blade) %>%
    st_collection_extract("LINESTRING") %>%
    rbind(network[-blade$nf, ])
}
```
`rowwise` from `dplyr` could have been used in place of the `purrr` functions here, though `map2` ended up being faster. 
```{r multiparts to singlparts, echo=F}
st_multipart_to_singleparts = function(x) {
  # Replicates Multipart to Singleparts algorithm in QGIS
  library(sf)
  
  mixed = st_sf(x)
  mixed$id = 1:nrow(mixed)
  geo_types = as.character(unique(st_geometry_type(mixed)))
  str_geo = grepl("MULTI", geo_types)
  
  if (sum(str_geo) == 0) {
    return(x)
  }
  
  multi = geo_types[str_geo]
  
  if (sum(str_geo) == length(geo_types)) {
    single = gsub("MULTI", "", multi)
  } else {
    single = geo_types[!str_geo]
  }
  
  multipart = mixed[st_geometry_type(mixed) == multi,]
  id = multipart$id
  rows = 1:nrow(multipart)
  
  singlepart = do.call(rbind, lapply(rows, function(s)
    suppressWarnings(st_cast(
      multipart[s,], single
    ))))
  
  col = colnames(mixed)
  rbind(mixed[-id, ], singlepart)[,col[col!="id"]]
}
```

```{r getting some data and splitting streets, message = F, warning =F,fig.height=6, fig.width=9}
library(sf)
library(dplyr)
library(tigris)
library(mapview)

addison = counties("vt", progress = F) %>% filter(COUNTYFP == "001") %>% st_transform(32145)

schools = read_sf(
  "https://opendata.arcgis.com/datasets/efa79839b0f841ac92c821a7b8afeda5_3.geojson"
) %>%
  st_transform(32145) %>%
  filter(lengths(st_intersects(., addison)) != 0)

streets = read_sf(
  "https://opendata.arcgis.com/datasets/1dee5cb935894f9abe1b8e7ccec1253e_39.geojson"
) %>%
  st_transform(32145) %>%
  st_intersection(st_geometry(addison)) %>%
  st_multipart_to_singleparts() %>% # need to make sure all features are linestrings  
  mutate(id = row_number())

blended = st_blend(schools, streets) %>%
  group_by(id) %>%
  mutate(cnt = n()) %>%
  ungroup

split_streets = filter(blended, cnt > 1) %>%
  mutate(color = ifelse(as.numeric(row.names(.)) %% 2 != 0, "split", "street"))

unsplit = filter(blended, cnt == 1)

mapview(
  select(split_streets, color),
  layer.name = "Split Streets",
  color = c("#F3D54E", "#69B3E7")
) +
  mapview(unsplit, layer.name = "Unsplit Streets", color = "#4C4B4C") +
  mapview(
    schools,
    layer.name = "Addison County Schools",
    col.regions = "#0D395F",
    color = "#000000",
    alpha = .2
  )
```

In this example, the linestrings nearest to the schools are either split or are left untouched if the nearest point from the school already exists as a vertex in the linestring. I found that `sfneworks` has a function called [`st_network_blend`](https://luukvdmeer.github.io/sfnetworks/reference/st_network_blend.html) after writing this so I may use that in the future, though writing `st_blend` was still a fun exercise.