---
title: "using dplyr and purrr in st_segments"
output: html_notebook
---

Although I was originally reluctant to solely use `dplyr` and `purrr` because my first two slower, I went on to try to replace both loops in [`st_segments`](https://kufreu.github.io/musings/st_segments/st_segments.nb.html) with `tidyverse` functions. 

```{r getting a road network, message=F, warning=F,echo =F}
library(sf)
library(dplyr)
library(tigris)

midd = read_sf("https://opendata.arcgis.com/datasets/625b92329bb841cf8295d14553b67e1f_2.geojson") %>% 
  filter(COLLEGE == "MIDDLEBURY COLLEGE") %>% 
  st_transform(32145) %>% # Vermont State Plane Meters
  st_union() %>% 
  st_centroid() %>% 
  st_buffer(5000) %>% 
  st_bbox %>% 
  st_as_sfc

streets = roads("vt", "addison county", progress_bar = F) %>% st_transform(32145) %>% st_intersection(midd) %>%
  filter(!grepl("MULTI", as.character(st_geometry_type(.))))
```


```{r st_segments_tbl}
st_segments_tbl = function(x, max_length) {
  # breaks linestrings or polygons into line segments, keeps attributes
  # x: layer with linestrings/polygons
  # max_length: optional parameter, max length of segments, unit from crs
  library(sf)
  library(dplyr)
  library(purrr)
  
  layer = mutate(x, value = row_number())
  
  if (!missing(max_length)) {
    layer = st_segmentize(layer, max_length)
  }
  
  coord = st_coordinates(layer)
  col = ncol(coord)
  
  geom = as_tibble(coord) %>%
    group_by(!!sym(colnames(coord)[col])) %>%
    mutate(X2 = lead(X),
           Y2 = lead(Y)) %>%
    ungroup() %>%
    filter(!is.na(X2)) %>%
    mutate(geometry = pmap(list(X, Y, X2, Y2), function(x, y, x2, y2)
      st_linestring(rbind(
        c(x, y), c(x2, y2)
      )))) %>%
    {
      st_as_sfc(.[["geometry"]], crs = st_crs(layer))
    }
  
  as_tibble(coord[, col]) %>%
    group_by(value) %>%
    slice(-1) %>%
    ungroup %>%
    left_join(st_drop_geometry(layer), by = "value") %>%
    select(-1) %>%
    mutate(geometry = geom) %>%
    st_sf()
}
```
Rather than casting the features to points and working with two geometry columns as I had with my first two attempts of the function, I went on to make the coordinate matrix from `st_coordinates` a tibble and replicated the loop in `st_segments` with `pmap` from `purrr` and `lead`. I also found that using `st_as_binary` here is wholly unnecessary and that doing without it leads to a considerable speed increase. Straying away from the loops also makes the function more readable. 
```{r base, echo = F}
st_segments = function(x, max_length) {
  # x: layer with linestrings/polygons
  # max_length: optional parameter, max length of segments, unit from crs
  library(sf)
  
  net = st_sf(x)
  net$fid = 1:nrow(net)
  
  if (!missing(max_length)) {
    net = st_segmentize(net, max_length)
  }
  
  coord = st_coordinates(net)
  col = ncol(coord)
  ids = coord[, col]
  feat = unique(ids)
  segments = list()
  
  for (i in feat) {
    coords = coord[ids == i,]
    for (j in 1:(nrow(coords) - 1)) {
      segments[[length(segments) + 1]] = st_as_binary(st_linestring(coords[j:(j + 1), 1:2]))
    }
  }
  
  fid = matrix(nrow = (length(ids) - length(feat)), ncol = 1)
  k = 0
  
  for (i in feat) {
    id = ids[ids == i]
    for (j in seq_along(id[1:(length(id) - 1)])) {
      fid[k + 1, ] = id[j]
      k = k + 1
    }
  }
  
  colnames(fid) = "fid"
  fid = as.data.frame(fid)
  fid = merge(fid,st_drop_geometry(net), all.x = T)[,-1, drop = F]
  fid$geometry = st_set_crs(st_as_sfc(segments), st_crs(net))
  st_sf(fid)
}
```


```{r benchmark, message = F, warning=F}
bench::mark(
  base = st_geometry(st_segments(streets, 1)),
  "dplyr/purrr" = st_geometry(st_segments_tbl(streets, 1)),
  iterations = 5
) %>%
  mutate(name = as.character(expression),
         max = bench::as_bench_time(lapply(time, max))) %>%
  select(
    name,
    min,
    median,
    max,
    "memory allocation" = mem_alloc,
    iterations = n_itr,
    "total time" = total_time
  )
```
`st_segments_tbl` uses significantly less memory than `st_segments` and takes a sixth of the time to run which is remarkable. It also runs faster than the previous `dplyr` version of `st_segments`. What really slows `st_segments` down seems to be the first loop as replacing the second loop with either `dplyr` or `data.table` code did allow for the function to run a little bit faster, though not considerably. This next benchmark shows that this is indeed the case. 

```{r loop and pmap, message=F,warning=F}
test = st_coordinates(streets)
bench::mark(
  loop = {
    (function(x) {
      col = ncol(x)
      ids = x[, col]
      feat = unique(ids)
      segments = list()

      for (i in feat) {
        coords = x[ids == i, ]
        for (j in 1:(nrow(coords) - 1)) {
          segments[[length(segments) + 1]] = st_as_binary(st_linestring(coords[j:(j + 1), 1:2]))
        }
      }
      st_as_sfc(segments)
    })(test)
  },
  tidy = {
    (function(x){
      as_tibble(x) %>%
        group_by(L1) %>%
        mutate(X2 = lead(X),
               Y2 = lead(Y)) %>%
        ungroup() %>%
        filter(!is.na(X2)) %>%
        mutate(geometry = pmap(list(X, Y, X2, Y2), function(x, y, x2, y2)
          st_linestring(rbind(
            c(x, y), c(x2, y2)
          )))) %>%
        {
          st_as_sfc(.$geometry)
        }
    })(test)
  },
  iterations = 50) %>%
  mutate(name = as.character(expression),
         max = bench::as_bench_time(lapply(time, max))) %>%
  select(
    name,
    min,
    median,
    max,
    "memory allocation" = mem_alloc,
    iterations = n_itr,
    "total time" = total_time
  )
```
I was having trouble using `sym` within `bench::mark` so I needed to explicitly state which column to group by in the tidy version. Even when working with a smaller dataset, the same difference in time can be observed between using a loop and `dplyr`/`purrr`. Looping through the `st_coordinates` matrix still takes around six times as long to run. Given just how much faster and more efficient `st_segments_tbl` is than `st_segments`,  `st_segments_tbl` may be my best option for now. Tibbles print nicely so that's also a plus.